---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# sapfluxnetr

[![Travis build status](https://travis-ci.org/sapfluxnet/sapfluxnetr.svg?branch=master)](https://travis-ci.org/sapfluxnet/sapfluxnetr)

`sapfluxnetr` provides tools for a tidy data analysis for the first global
database of sap flow measurements
([Sapfluxnet Project](http://sapfluxnet.creaf.cat))

## Examples

You can work with individual sites:

```{r example_single}
# load packages
library(sapfluxnetr)
# install.packages("tidyverse")
library(tidyverse)

# ARG_MAZ example site data
data('ARG_MAZ', package = 'sapfluxnetr')
data('sfn_metadata_ex', package = 'sapfluxnetr')

# plot site sapflow measurements versus vpd
sfn_plot(ARG_MAZ, formula_env = ~ vpd)

# daily sapflow and environmental metrics
arg_maz_metrics <- daily_metrics(
  ARG_MAZ, tidy = TRUE, metadata = sfn_metadata_ex
)

# plot daily aggregations
ggplot(arg_maz_metrics, aes(x = vpd_q_95, y = sapflow_q_95, colour = pl_code)) +
  geom_point()
```

And you can work with multiple sites:

```{r example_multi}
# ARG_TRE and AUS_CAN_ST2_MIX example sites
data('ARG_TRE', package = 'sapfluxnetr')
data('AUS_CAN_ST2_MIX', package = 'sapfluxnetr')
multi_sfn <- sfn_data_multi(ARG_TRE, ARG_MAZ, AUS_CAN_ST2_MIX)

# plotting the individual sites. It creates a list of plots
plots_list <- sfn_plot(multi_sfn, formula_env = ~ vpd)
plots_list[['AUS_CAN_ST2_MIX']]

# daily sapflow standard metrics
multi_metrics <- daily_metrics(
  multi_sfn, tidy = TRUE, metadata = sfn_metadata_ex
)

# plot daily aggregations
ggplot(multi_metrics, aes(x = vpd_q_95, y = sapflow_q_95, colour = si_code)) +
  geom_point(alpha = 0.2)
```

## Installation

You can install sapfluxnetr from CRAN:

```{r cran-installation, eval = FALSE}
install.packages('sapfluxnetr')
```

Be advised, `sapfluxnetr` is in active development and can contain undiscovered
bugs. If you find something not working as expected fill a bug at
https://github.com/sapfluxnet/sapfluxnetr/issues

## Overview

Please see `vignette('sapfluxnetr-quick-guide', package = 'sapfluxnetr')` for
a detailed overview of the package capabilities.