---
title: "sapfluxnetr Quick Guide"
author: "Víctor Granda (Sapfluxnet Team)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sapfluxnetr Quick Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`sapfluxnetr` R package provides tools for a tidy data analysis for the first
sap flow measurements global database 
([Sapfluxnet Project](http://sapfluxnet.creaf.cat/app)). In this vignette you
will learn how to install the package, download the data and get started with
some data metrics.

## Installing the package

`sapfluxnetr` stable versions are in the CRAN repository, so the installation of
the package is as per usual:

```{r cran_inst, eval=FALSE}
install.packages('sapfluxnetr')
```

Development versions of the package reside in github. If you want the lastest
updates (and also the latests bugs, be advised ;) ), please install the devel
branch of the [github repository](https://github.com/sapfluxnet/sapfluxnetr)
with the `devtools` package:

```{r github_inst, eval=FALSE}
devtools::install_github('sapfluxnet/sapfluxnetr', ref = 'devel')
```

Now you can load the package, and also the `tidyverse` package, as it will be
needed later:

```{r pkg_load}
library(sapfluxnetr)
library(tidyverse)
```


## Download the data from Zenodo

Sapfluxnet data can is hosted in Zenodo, and `sapfluxnetr` provides functions
for downloading the data directly from R (for sapfluxnet data contributors at
the moment, after one embargo year, the data will be available for anyone):

>TODO: zenodo's part is still in development, nothing to see here, please continue

## Inspecting a site

We have downloaded the data, *and now, what?*.  
First, let's get used to the data structure that Sapfluxnet Data provides, and
for that we will choose a site and start playing with it. In this example we
will use the `ARG_MAZ` site, as it is small and it will be fast to see the
package capabilities. There are sites like `FRA_PUE` 20 times bigger than this,
but as you can imagine, the time is also increasing when analising those
bigger datasets.  
So, let's read the data in the environment:

```{r read_and_inspect, eval=FALSE}
# read the data
arg_maz <- read_sfn_data('ARG_MAZ', folder = 'data/plant')

# see a brief summary of the site:
arg_maz
```

```{r read_internals, echo=FALSE}
# read the data
arg_maz <- read_sfn_data('ARG_MAZ', folder = '../tests/testthat/big_test/plant')

# see a brief summary of the site:
arg_maz
```


At first glance, we know by this summary that is an Argentinian forest (first
three letters of the site code are always the country code),
contributed by Sebastian Pfautsch and Pablo Peri with 5 *Nothofagus pumilio*
trees measured along 15 days in 2009. Also we can see the environmental variables
measured (`ta`, `rh`, `vpd`, `sw_in`, `ws`, `precip`, `swc_shallow`, `ppfd_in`
and `ext_rad`) and the biome classification. Finally, we can see that the
environmental data has some flags (more on that later).

That's nice, *but where is the data?*.  
`sfn_class` objects have different slots containing the different data, each of
one has an accesor function (see `?sfn_get_methods` for detailed info):

```{r accesors_data}
# sapf data with original site timestamp
arg_maz_sapf <- get_sapf_data(arg_maz, solar = FALSE)
arg_maz_sapf

# env_data with calculated aparent solar time
arg_maz_env <- get_env_data(arg_maz, solar = TRUE)
arg_maz_env
```

You can see that the TIMESTAMP variable changes between both kinds of data. That
is because the TIMESTAMP returned is controled by the `solar` parameter
(see `?sfn_get_methods`).

Metadata can be accesed in the same way:

```{r accesors_md}
arg_maz_site_md <- get_site_md(arg_maz)
arg_maz_site_md
arg_maz_stand_md <- get_stand_md(arg_maz)
arg_maz_stand_md
arg_maz_species_md <- get_species_md(arg_maz)
arg_maz_species_md
arg_maz_plant_md <- get_plant_md(arg_maz)
arg_maz_plant_md
arg_maz_env_md <- get_env_md(arg_maz)
arg_maz_env_md
```

If in doubt about some of the metadata variables (what it means, units...) a
description can be obtained from `describe_md_variable` function:

```{r describe}
# what is env_ta?
describe_md_variable('env_ta')
# or pl_species?
describe_md_variable('pl_species')
```

We can also plot the different data with the help of `sfn_plot` function. It will
return `ggplot` objects that can be modified afterwards:

```{r sapf_plot}
sfn_plot(arg_maz, type = 'sapf', solar = TRUE) +
  facet_wrap(~ Tree) + theme(legend.position = 'none')
```

```{r env_plot}
sfn_plot(arg_maz, type = 'env', solar = TRUE) +
  facet_wrap(~ Variable, scales = 'free_y') + theme(legend.position = 'none')
```

We can also plot environmental variables individually,

```{r vpd}
sfn_plot(arg_maz, type = 'vpd', solar = TRUE)
```

or an environmental variable *versus* the sap flow measurements

```{r vs_sapf}
sfn_plot(arg_maz, formula = ~vpd, solar = TRUE) +
  facet_wrap(~ Tree) + theme(legend.position = 'none')
```




## Preparing the metadata

Sapfluxnet metadata is stored along the 


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
