---
title: "Custom aggregation"
author: "Victor Granda (Sapfluxnet Team)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`sapfluxnetr` package offers a very flexible system based on `tidyverse` and
`tibbletime` packages to aggregate and summarise the site/s data in the form
of the `sfn_metrics` functions. All the metrics family of functions (`?metrics`)
make use of `sfn_metrics` under the hood. If you want full control to the
metrics returned and temporal periods to aggregate we recommend you to use this
function. This vignette will show you how.

## Custom summarising functions

`daily_metrics` and related functions return a complete set of metrics ready for
use, but if you want simpler metrics and/or avoid the computational burden of
returning all the pre-fixed metrics you can supply your own summarising metrics:

```{r}
# libraries
library(sapfluxnetr)
library(dplyr)

### only mean and sd at a daily scale
# data
data('FOO', package = 'sapfluxnetr')
# summarising funs (built with funs function in dplyr package)
custom_funs <- funs(mean = mean(., na.rm = TRUE), std_dev = sd(., na.rm = TRUE))
# metrics
foo_simpler_metrics <- sfn_metrics(
  FOO,
  period = 'daily',
  .funs = custom_funs,
  solar = TRUE,
  general = TRUE,
  predawn = TRUE, pd_start = 5, pd_end = 7,
  midday = TRUE, md_start = 11, md_end = 13,
  nighttime = FALSE
)

foo_simpler_metrics[['sapf']][['sapf_gen']]
```

You can also select if the "special interest" intervals (predawn, midday or
nighttime) are calculated or not. For example, if you are only interested in the
midday interval you can use:

```{r}
foo_simpler_metrics_midday <- sfn_metrics(
  FOO,
  period = 'daily',
  .funs = custom_funs,
  solar = TRUE,
  general = FALSE,
  predawn = FALSE,
  midday = TRUE, md_start = 11, md_end = 13,
  nighttime = FALSE
)

foo_simpler_metrics_midday[['sapf']][['sapf_md']]
```

## Custom aggregation periods

`period` argument in `sfn_metrics` is passed to `collapse_index` function in the
`tibbletime` package, and so, it can use the same input:

  + "frequency period" format, where frequency is a number and period is an
    interval as character (i.e. "1 year" same as "yearly", "7 days")

```{r}
# weekly
foo_weekly <- sfn_metrics(
  FOO,
  period = '7 days',
  .funs = custom_funs,
  solar = TRUE,
  general = FALSE,
  predawn = FALSE,
  midday = TRUE, md_start = 11, md_end = 13,
  nighttime = FALSE
)

foo_weekly[['env']][['env_md']]
```

 + A vector of dates as period boundaries. This way you can build irregular
   periods

```{r}
# custom
foo_custom <- sfn_metrics(
  FOO,
  period = as.POSIXct(
    c('2009-11-17 00:00:00', '2009-11-22 14:00:00', '2009-11-30 23:59:00'),
                      tz = 'UTC'
  ),
  .funs = custom_funs,
  solar = TRUE,
  general = FALSE,
  predawn = FALSE,
  midday = TRUE, md_start = 11, md_end = 13,
  nighttime = FALSE
)
foo_custom[['sapf']][['sapf_md']]
```

