---
title: "Custom aggregation"
author: "Victor Granda (Sapfluxnet Team)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom aggregation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`sapfluxnetr` package offers a very flexible but powerfull system based on
`tidyverse` and `tibbletime` packages to aggregate and summarise the site/s data
in the form of the `sfn_metrics` functions. All the metrics family of functions
(`?metrics`) make use of `sfn_metrics` under the hood. If you want full control
to the metrics returned and temporal periods to aggregate we recommend you to
use this main function. This vignette will show you how.

## Pre-fixed summarising functions

1. `daily_metrics`
1. `monthly_metrics`
1. `predawn_metrics`
1. `midday_metrics`
1. `nightly_metrics`
1. `daylight_metrics`

See each function help for a detailed description and examples of use.

## Custom summarising functions

`daily_metrics` and related functions return a complete set of metrics ready for
use, but if you want simpler metrics and/or avoid the computational burden of
returning all the pre-fixed metrics you can supply your own summarising metrics
using the `funs` function belonging to the `dplyr` package in the `tidyverse`
suite:

```{r}
# libraries
library(sapfluxnetr)
library(dplyr)

### only mean and sd at a daily scale
# data
data('ARG_TRE', package = 'sapfluxnetr')

# summarising funs (built with funs function in dplyr package)
custom_funs <- funs(mean = mean(., na.rm = TRUE), std_dev = sd(., na.rm = TRUE))

# metrics
foo_simpler_metrics <- sfn_metrics(
  ARG_TRE,
  period = 'daily',
  .funs = custom_funs,
  solar = TRUE,
  interval = 'general'
)

foo_simpler_metrics[['sapf']]
```

You can also select if the "special interest" intervals (predawn, midday,
nighttime or daylight) are calculated or not. For example, if you are only
interested in the midday interval you can use:

```{r}
foo_simpler_metrics_midday <- sfn_metrics(
  ARG_TRE,
  period = 'daily',
  .funs = custom_funs,
  solar = TRUE,
  interval = 'midday', int_start = 11, int_end = 13
)

foo_simpler_metrics_midday[['sapf']]
```

> When supplying only one function names of variables are not changed to
  contain the metric name at the end, as the summary function returns the
  same columns as the original data

## Custom aggregation periods

`period` argument in `sfn_metrics` is passed to `collapse_index` function in the
`tibbletime` package, and so, it can use the same input:

  + "frequency period" format, where frequency is a number and period is an
    interval as character (i.e. "1 year" same as "yearly", "7 days")

```{r}
# weekly
foo_weekly <- sfn_metrics(
  ARG_TRE,
  period = '7 days',
  .funs = custom_funs,
  solar = TRUE,
  interval = 'general'
)

foo_weekly[['env']]
```

 + A vector of dates as period boundaries. This way you can build irregular
   or custom periods.

```{r}
# custom
foo_custom <- sfn_metrics(
  ARG_TRE,
  period = as.POSIXct(
    c('2009-11-17 00:00:00', '2009-11-22 14:00:00', '2009-11-30 23:59:00'),
                      tz = 'UTC'
  ),
  .funs = custom_funs,
  solar = TRUE,
  interval = 'general'
)
foo_custom[['sapf']]
```

## Extra parameters

`sfn_metrics` has a `...` parameter intended to supply additional parameters to
the internal functions used:

  1. `tibbletime::collapse_index` accepts the following extra arguments:
     
     - `start_date`  
        NULL by default in the sfn_metrics implementation
     - `side`  
        "start" by default in the sfn_metrics implementation
     - `clean`  
        TRUE by default in the sfn_metrics implementation
  
  1. `dplyr::summarise_all` accepts extra arguments intended to be applied to
     summarising functions provided (to all, so they all must have the argument
     provided or an error will be raised). That's the reason because we
     recommend to use the `funs` function from `dplyr`, as the arguments are
     specified for the individual functions.

For example, if we want the TIMESTAMPs after aggregation to show the end of the
period instead the beggining (default) we can do the following:

```{r}
foo_simpler_metrics_end <- sfn_metrics(
  ARG_TRE,
  period = 'daily',
  .funs = custom_funs,
  solar = TRUE,
  interval = 'general',
  side = "end"
)

foo_simpler_metrics_end[['sapf']]
```

If it is compared with the `foo_simpler_metrics` calculated before, now the
period is identified in the TIMESTAMP by the ending of the period (daily in this
case).
