---
title: "Metadata and Data units"
author: "Victor Granda (Sapfluxnet Team)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Metadata and Data units}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Metadata

Metadata numeric variables come in different units depending on the variable.
One easy way to check for a variable units is to use the `describe_md_variable`
function:

```{r describe_md_variable}
library(sapfluxnetr)
describe_md_variable('si_elev')
describe_md_variable('st_age')
```

As it can be seen, site elevation is in meters and stand age is in years.

Nevertheless, and for the sake of completion, here you can find a complete list
of the metadata variables and their units:

```{r md_vars_table, echo=FALSE, results='asis'}
suppressMessages(library(dplyr))
library(magrittr)
site_md_table <- sapfluxnetr:::.metadata_architecture() %>%
  magrittr::extract2(., 'site_md') %>%
  purrr::map_dfr(magrittr::extract, c('description', 'type', 'units')) %>%
  dplyr::mutate(
    variable = sapfluxnetr:::.metadata_architecture() %>%
      magrittr::extract2(., 'site_md') %>%
      names()
  ) %>%
  select(variable, everything())

stand_md_table <- sapfluxnetr:::.metadata_architecture() %>%
  magrittr::extract2(., 'stand_md') %>%
  purrr::map_dfr(magrittr::extract, c('description', 'type', 'units')) %>%
  dplyr::mutate(
    variable = sapfluxnetr:::.metadata_architecture() %>%
      magrittr::extract2(., 'stand_md') %>%
      names()
  ) %>%
  select(variable, everything())

species_md_table <- sapfluxnetr:::.metadata_architecture() %>%
  magrittr::extract2(., 'species_md') %>%
  purrr::map_dfr(magrittr::extract, c('description', 'type', 'units')) %>%
  dplyr::mutate(
    variable = sapfluxnetr:::.metadata_architecture() %>%
      magrittr::extract2(., 'species_md') %>%
      names()
  ) %>%
  select(variable, everything())

plant_md_table <- sapfluxnetr:::.metadata_architecture() %>%
  magrittr::extract2(., 'plant_md') %>%
  purrr::map_dfr(magrittr::extract, c('description', 'type', 'units')) %>%
  dplyr::mutate(
    variable = sapfluxnetr:::.metadata_architecture() %>%
      magrittr::extract2(., 'plant_md') %>%
      names()
  ) %>%
  select(variable, everything())

env_md_table <- sapfluxnetr:::.metadata_architecture() %>%
  magrittr::extract2(., 'env_md') %>%
  purrr::map_dfr(magrittr::extract, c('description', 'type', 'units')) %>%
  dplyr::mutate(
    variable = sapfluxnetr:::.metadata_architecture() %>%
      magrittr::extract2(., 'env_md') %>%
      names()
  ) %>%
  select(variable, everything())

bind_rows(
  site_md_table, stand_md_table, species_md_table, plant_md_table, env_md_table
) %>%
  xtable::xtable(align = c('lcccc')) %>%
  print(type = 'html')
```

## Data

### Sap flow units

Sap flow units depends on the database level chosen:

$$Plant:~\frac{cm³}{s}$$
$$Sapwood:~\frac{cm³}{cm²_{sa}·s}$$
$$Leaf:~\frac{cm³}{cm²_{la}·s}$$

### Environmental variables

Environmental variables come in the following units:

```{r environmetal_vars_table, echo=FALSE, results='asis'}
tibble::tibble(
  Variable = c(
    'env_ta', 'env_rh', 'env_vpd', 'env_sw_in', 'env_ppfd', 'env_netrad',
    'env_ws', 'env_precip', 'env_swc_shallow', 'env_swc_deep'
  ),
  Description = c(
    'Air temperature', 'Air relative humidity',
    'Vapour pressure deficit', 'Shortwave incoming radiation',
    'Incoming photosynthetic photon flux density',
    'Net radiation', 'Wind speed', 'Precipitation',
    'Shallow soil water content',
    'Deep soil water content'
  ),
  Units = c(
    'ºC', '%', 'kPa', 'W m-2', 'micromols m-2 s-1', 'W m-2', 'm s-1', 'mm timestep-1',
    'cm3 cm-3', 'cm3 cm-3'
  )
) %>%
  xtable::xtable(align = c('lccc')) %>%
  print(type = 'html')
```

## Timestamps

### TIMESTAMP

`TIMESTAMP` variable comes as provided by the contributors and timezone is
applied per the site metadata description:

```{r TIMESTAMP_var}
library(dplyr)
library(lubridate)

# timezone provided by contributor
get_env_md(ARG_TRE) %>% pull(env_time_zone)

# timezone in the TIMESTAMP
get_timestamp(ARG_TRE) %>% tz()
```

As you can see, the final timezone is in GMT format.

### Solar TIMESTAMP

When inspecting the solar `TIMESTAMP` variable, timezone declared is UTC:

```{r solar_TIMESTAMP}
get_solar_timestamp(ARG_TRE) %>% tz()
```

This is so because solar timestamp is only refered to the sun position in the
sky, rendering the timezone useless in this case. So UTC was chosen as timezone
to be able to join and compare global data.
